Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-23 text/ms.tex
More intro tweaks

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, but
systematic evidence has been elusive. A survey of ecologists indicated that
surprising results in field experiments may be far more common than we assume
\citep{doak2008}. And, a number of major environmental events on geologic and
contemporary time scales could be considered to have caused ecological black
swans \citep{nunez2012}. Other work has detected heavy-tail dynamics for
particular systems. For example, \citet{segura2013} identified heavy tails in
time series of marine microbe populations and \citet{drake2014} identified
heavy tails in time to extinction for experimental waterflea populations.
However, systematic evidence has been limited. What appeared to be heavy tails
in breeding bird abundance \citep{keitt1998}, could also be explained by the
mixing of species in the original dataset \citep{allen2001}. A subsequent
large scale analysis of the Global Population Dynamics Database (GPDD) found
little evidence for heavy-tail distributions in longer abundance time series
\citet{halley2002}.

@@

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population
dynamics are not heavy tailed (e.g.~REF, REF, REF).
Alternatively, dynamics might be heavy tailed, but a paucity of statistical
methods appropriate for short time series has limited our
ability to detect them.
For example, (1) the generalized extreme value (GEV) distribution has been
applied to environmental data (REF).
The GEV describes the distribution of the most extreme
event per time interval (e.g.~heaviest rainfall per year), but this approach
requires long time series that can be condensed into time intervals (REF).
(2) Heavy- and thin-tailed distributions have been competed with each
other (REF), but information criteria likely have
trouble accurately distinguishing such subtle differences in population
dynamics for short time series (REF).
(3) A catastrophic mixture distribution model allows one to elegantly quantify
the probability that population events are catastrophic is also data intensive
\citep{ward2007}.
(4) Other research has focussed on tipping points and regime shifts, but such
methods are usually focussed on entire communities or ecosystems (e.g.\ REF).
Furthermore, black swans are more general than tipping points --- they don't
necessarily imply a permanent change of state.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model (REFs?). The Gompertz model represents population growth as a linear
function in $\log$ space. If we let $x_t$ represent the $\log$ abundance ($N$)
at time $t$, we can represent the Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious model
\citep{brook2006}.

>>>>

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, but
systematic evidence has been elusive. A survey of ecologists indicated that
surprising results in field experiments may be far more common than we assume
\citep{doak2008}, and a number of major environmental events on geologic and
contemporary time scales could be considered to have caused ecological black
swans \citep{nunez2012}. Other work has detected heavy-tail dynamics for
particular systems. For example, \citet{segura2013} identified heavy tails in
time series of marine microbe populations and \citet{drake2014} identified
heavy tails in time to extinction for experimental waterflea populations.
However, systematic evidence has been limited. What appeared to be heavy tails
in breeding bird abundance \citep{keitt1998}, could also be explained by the
mixing of species in the original dataset \citep{allen2001}. A subsequent
large scale analysis of the Global Population Dynamics Database (GPDD) found
little evidence for heavy-tail distributions in longer abundance time series
\citet{halley2002}.

@@

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population
dynamics are not heavy tailed (e.g.~REF, REF, REF).
Alternatively, dynamics might be heavy tailed, but a paucity of statistical
methods appropriate for short time series has limited our
ability to detect them.
As four examples: (1) The generalized extreme value (GEV) distribution has been
applied to environmental data (REF).
The GEV describes the distribution of the most extreme
event per time interval (e.g.~heaviest rainfall per year), but this approach
requires long time series that can be condensed into time intervals (REF).
(2) Heavy- and thin-tailed distributions have been competed with each
other (REF), but information criteria likely have
trouble accurately distinguishing such subtle differences in population
dynamics for short time series (REF).
(3) A catastrophic mixture distribution model allows one to quantify the
probability that population events are catastrophic but is also relatively data
intensive \citep{ward2007}.
(4) Other research has focussed on tipping points and regime shifts, but such
methods are usually focussed on entire communities or ecosystems (e.g.\ REF).
Furthermore, black swans are more general than tipping points --- they don't
necessarily imply a permanent change of state.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model (REFs?). The Gompertz model represents population growth as a linear
function in $\log$ space. If we let $x_t$ represent the $\log$ abundance at
time $t$, we can represent the Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious model
\citep{brook2006}.
